Interactive effects of drought and habitat fragmentation on vital rates of an understory tropical plant

### 2020-12-03 analysis/paper/paper.Rmd
edited citations

# Methods
## Study site

The Biological Dynamics of Forest Fragments Project (BDFFP) is located ~70 km north of Manaus, Brazil (2º30' S, 60ºW). In addition to large expanses of continuous forest, the BDFFP has forest fragment reserves ranging in size from 1-100 ha (Fig. --). These fragments were isolated from 1980-1984 by felling the trees surrounding the forest patch chosen for isolation and, in most cases, burning the downed trees once they dried. The vegetation regenerating around fragments is cleared periodically to ensure their isolation (Bierregaard et al. 2001). The BDFFP reserves are located in nonflooded (i.e., *terra firme*) tropical lowland forest with a 30-37m tall canopy, emergent trees up to 55m (Rankin-de Merona et al. 1992), and an understory dominated by stemless palms (Scariot 1999). The soils in the reserves are nutrient-poor xanthic ferrosols with poor water retention capacity despite their high clay content. Mean annual temperature in the region is 26º C (range=19–39º C), and annual rainfall at the BDFFP reserves ranges from 1900-2300 mm with a pronounced dry season from June to October (Appendix?).

## Focal species 
*Heliconia acuminata* (LC Rich.) (Heliconiaceae) is a perennial monocot widely distributed throughout the Central Amazon (Kress 1990), and it is the most abundant understory herb at the BDFFP (Ribeiro et al. 2010). While many *Heliconia* species grow in large patches along forest edges or in disturbed areas, others -- such as *H. acuminata* -- are found at lower densities in shaded forest understories. Species in this group produce fewer infloresences and are pollinated by traplining hummingbirds that move from one plant to the next rather than defend a territory (Berry and Kress 1991). In our sites *H. acuminata* is pollinated by *Phaeothornis superciliosus* and *P. bourcieri* (Bruna et al. 2004); reproducctive plants will have 1-X flowering shoots , wach of which has an inflorescence with 20--25 flowers [@bruna2002]. Fruits mature April-May and are consumed by a thrush and several species of manakin (Uriarte et al. 2011). Dispersed seeds (1-3 seeds per fruit,	$\bar{x}$=2) germinate ~6 months after dispersal at the beginning of the rainy season, and have higher rates of germination and seedling establishment in continuous forest than forest fragments (Bruna 1999, 2002). *Heliconia acuminata* is self-incompatible and does not reproduce vegetatively.

## Demographic data collection
In 1997-1998 we established a series of 5000 $m^{2}$ plots ($50 \times 100$m) in the BDFFP reserves in which we marked and measured all *H. acuminata*. The plots are located in continuous forest (n=6), 10-ha fragments (n=3), and 1-ha fragments (n=4), with distance between plots ranging from 500 m-41 km. Plots in 1-ha fragments were on one randomly selected half of the fragment, plots in 10-ha fragments were in the fragment center, and plots in continuous forest were placed in locations 500-4000 m from the borders of secondary and mature forest (Fig or Table). Each plot is subdivided into 50 quadrats ($10 \times 10$m) to simplify surveys, during which we recorded the identity of flowering plants, the number of vegetative shoots each plant had, and the height of each plant to the tallest leaf. Height and shoot number are correlated with leaf area, the probability of flowering, and rates of survivorship (Bruna and Kress 2002, Bruna et al. 2002). In 2008 we mapped the location of all established plants to the nearest 10 cm; in subsequent years we did the same with all new seedlings. There is a seven-fold range in H. acuminata density in the forest plots, with density in several of the plots comparable to that in low-density fragments (Bruna and Kress 2002).

>>>>

# Methods

## Study site

The Biological Dynamics of Forest Fragments Project (BDFFP) is located \~70 km north of Manaus, Brazil (2º30' S, 60ºW). In addition to large expanses of continuous forest, the BDFFP has forest fragment reserves ranging in size from 1-100 ha (Fig. --). These fragments were isolated from 1980-1984 by felling the trees surrounding the forest patch chosen for isolation and, in most cases, burning the downed trees once they dried. The vegetation regenerating around fragments is cleared periodically to ensure their isolation (Bierregaard et al. 2001). The BDFFP reserves are located in nonflooded (i.e., *terra firme*) tropical lowland forest with a 30-37m tall canopy, emergent trees up to 55m (Rankin-de Merona et al. 1992), and an understory dominated by stemless palms (Scariot 1999). The soils in the reserves are nutrient-poor xanthic ferrosols with poor water retention capacity despite their high clay content. Mean annual temperature in the region is 26º C (range=19--39º C), and annual rainfall at the BDFFP reserves ranges from 1900-2300 mm with a pronounced dry season from June to October (Appendix?).

## Focal species

*Heliconia acuminata* (LC Rich.) (Heliconiaceae) is a perennial monocot widely distributed throughout the Central Amazon [@kressDiversityDistributionHeliconia1990], and it is the most abundant understory herb at the BDFFP [@ribeiroInfluencePostclearingTreatment2010]. While many *Heliconia* species grow in large patches along forest edges or in disturbed areas, others -- such as *H. acuminata* -- are found at lower densities in shaded forest understories. Species in this group produce fewer infloresences and are pollinated by traplining hummingbirds that move from one plant to the next rather than defend a territory [@berryHeliconiaIdentificationGuide1991a]. In our sites *H. acuminata* is pollinated by *Phaeothornis superciliosus* and *P. bourcieri* [@brunaHeliconiaAcuminataReproductive2004]; reproductive plants will have 1-X flowering shoots , wach of which has an inflorescence with 20--25 flowers [@brunaHabitatFragmentationDemographic2002]. Fruits mature April-May and are consumed by a thrush and several species of manakin [@uriarteDisentanglingDriversReduced2011]. Dispersed seeds (1-3 seeds per fruit, $\bar{x}$=2) germinate \~6 months after dispersal at the beginning of the rainy season, and have higher rates of germination and seedling establishment in continuous forest than forest fragments [@brunaHabitatFragmentationDemographic2002; @brunaSeedGerminationRainforest1999]. *Heliconia acuminata* is self-incompatible and does not reproduce vegetatively.

## Demographic data collection

In 1997-1998 we established a series of 5000 $m^{2}$ plots ($50 \times 100$m) in the BDFFP reserves in which we marked and measured all *H. acuminata*. The plots are located in continuous forest (n=6), 10-ha fragments (n=3), and 1-ha fragments (n=4), with distance between plots ranging from 500 m-41 km. Plots in 1-ha fragments were on one randomly selected half of the fragment, plots in 10-ha fragments were in the fragment center, and plots in continuous forest were placed in locations 500-4000 m from the borders of secondary and mature forest (Fig or Table). Each plot is subdivided into 50 quadrats ($10 \times 10$m) to simplify surveys, during which we recorded the identity of flowering plants, the number of vegetative shoots each plant had, and the height of each plant to the tallest leaf. Height and shoot number are correlated with leaf area, the probability of flowering, and rates of survivorship [@brunaHabitatFragmentationDemographic2002; @brunaEffectsForestFragmentation2002]. In 2008 we mapped the location of all established plants to the nearest 10 cm; in subsequent years we did the same with all new seedlings. There is a seven-fold range in H. acuminata density in the forest plots, with density in several of the plots comparable to that in low-density fragments [@brunaHabitatFragmentationDemographic2002].
